NS notes

### 2012-06-06 docs/NavierStokes/notes.tex
adding figs and NS notes

\newcommand{\vect}[1]{\ensuremath\boldsymbol{#1}}
\newcommand{\tensor}[1]{\underline{\vect{#1}}}
\newcommand{\del}{\triangle}
\newcommand{\grad}{\nabla}
\renewcommand{\div}{\grad \cdot}
\newcommand{\ip}[1]{\left\langle #1 \right\rangle}
\newcommand{\eip}[1]{a\left( #1 \right)}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdd}[2]{\frac{\partial^2#1}{\partial#2^2}}

@@

@@

\section{Intro}

Following Demkowicz and Oden, we seek to solve 
\[
\sum^2_{i=1} G_i\left( U\right)_{,i}=\sum^2_{i=1} V_i\left( U\right)_{,i}
\]
where
\[
V_i = \vectfour{0}{\sigma_{i1}}{\sigma_{i2}}{\sum_{j=1}^2\sigma_{ij}u_j + q_i}
\]
are the viscous fluxes, and 
\[
 G_1 = \vectfour{\rho u_1}{\rho u_1^2+p}{\rho u_1 u_2}{((\rho e)+p)u_1}, \qquad G_2 = \vectfour{\rho u_2}{\rho u_1 u_2}{\rho u_2^2+p}{((\rho e)+p)u_2}
\]
are the Eulerian fluxes.

We also have the constitutive relations 
\begin{align*}
\sigma_{ij} &= \mu(u_{i,j}+u_{j,i}) + \lambda u_{k,k}\delta_{ij}\\
q_i &= \kappa T_{,i}\\
p &= (\gamma-1)\rho\iota\\
\iota &= e-\frac{1}{2}(u_1^2+u_2^2)\\
\iota &= c_vT
\end{align*}
where $e$ and $\iota$ are energy and internal energy per unit mass, respectively.

\section{Symbolic form}
In symbolic form, we can write the Navier Stokes equations as a system of seven first order equations. Let $[u_1,u_2] = [u,v]$.

@@

In general, $\mu$ and $\lambda$ are functions of temperature.

\subsection{Fourier's heat conduction law}
\begin{align*}
\vec{q} &= \kappa \grad T
\end{align*}
$\kappa$ is generally a function of temperature, and is often assumed to be proportional to $\mu$.

We introduce here the Prandtl number here as well
\[
{\rm Pr} = \frac{\gamma c_v \mu}{\kappa}
\]
In this case, we assume a constant Prandlt number, implying that viscosity is proportional to conductivity.

@@

We strongly enforce symmetry of $\boldsymbol \sigma$ by setting $\sigma_{21} = \sigma_{12}$.

\section{Nondimensionalization}
To nondimensionalize our equations, we introduce nondimensional quantities for length, density, velocity, temperature, and viscosity. 
\[
\boldsymbol x^* = \frac{\boldsymbol x}{L}, \qquad \rho^* = \frac{\rho}{\rho_{\infty}}, \qquad \boldsymbol u^* = \frac{\boldsymbol u}{V_\infty}, \qquad T^* = \frac{T}{T_\infty}, \qquad \mu^* = \frac{\mu}{\mu_\infty}
\]
Momentum, pressure, internal energy, and bulk viscosity are then nondimensionalized with respect to the above variables
\[
\boldsymbol m^* = \frac{\boldsymbol m}{\rho_\infty V_\infty}, \qquad p^* = \frac{p}{\rho_\infty V_\infty^2}, \qquad \iota^* = \frac{\iota}{V_\infty^2}, \qquad \lambda^* = \frac{\lambda}{\mu_\infty}
\]
We introduce, for convenience, the Reynolds number
\[
{\rm Re} = \frac{\rho_\infty V_\infty L}{\mu_\infty} 
\]
and the reference (free stream) Mach number
\[
M_\infty = \frac{V_\infty}{\sqrt{\gamma(\gamma-1)c_vT_\infty}}
\]
Note that 
\[
a = \sqrt{\frac{\gamma p_\infty}{\rho_\infty}} = \sqrt{{\gamma p_\infty}} = \sqrt{\gamma(\gamma-1)c_vT_\infty}
\]
The equations take the same form as before after nondimensionalization, so long as we define new material constants
\[
\tilde{\mu} = \frac{\mu^*}{\rm Re}, \qquad \tilde{\lambda} = \frac{\lambda^*}{\rm Re}, \qquad \tilde{c}_v = \frac{1}{\gamma(\gamma-1)M_\infty^2}, \qquad \tilde{\kappa} = \frac{\gamma\tilde{c}_v\tilde{\mu}}{\rm Pr}
\]
From here on, we drop the $^*$ superscript and adopt the nondimensionalized equations.

@@

@@

For our initial guesses, we set
\[
\rho = 1,\qquad m_1= 1,\qquad m_2 = 0, \qquad T = 1
\]
which is consistent with what was done in Demkowicz and Oden (find citation).

\section{	Boundary conditions}

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=.5]{flat_plate_BCs.pdf}
\caption{Carter flat plate problem.}
\end{figure}

\begin{itemize}
\item \textbf{Inflow boundary conditions:} free stream conditions are applied here to all four fluxes $\widehat{f}_{i,n}$.
\item \textbf{Symmetry boundary conditions:} $u_n = q_n = \pd{u_s}{n} = 0$. Here, this implies $u_2 = q_2 = \sigma_{12} = 0$. We impose the stress condition by noting that, for the flat plate geometry, if $u_2 = 0$, then at the top and bottom, with $n = (0,1)$, $\widehat{f}_{2,n} = \sigma_{12}$, and $\widehat{f}_{4,n} = q_2$ if $\sigma_{12}$ and $u_2 = 0$. 
\item \textbf{Flat plate boundary conditions:} $u_1 = u_2 = 0$, and $T = T_w = \left[1+(\gamma-1)M_\infty^2/2\right] T_\infty = 2.8T_\infty$ (for Mach 3 flow). We impose these strongly on the trace variables $\widehat{u}_1, \widehat{u}_2, \widehat{T}$. 
\item \textbf{Outflow boundary conditions:} the exact boundary conditions to enforce here are questionable\footnote{Demkowicz et al enforce this boundary condition only in regions where the flow is subsonic.}. Capon et al enforce $\pd{u_1}{n}=\pd{u_2}{n}=0$ and $\pd{T}{n} = 0$. Since $u_1$ and $u_2$ are not expected to be zero at these points, we will set $\widehat{f}_{2,n}$ and $\widehat{f}_{4,n}$ to their represented quantities using either the field variables from the background flow, or (in the case of pseudo-time stepping) the previous timestep's flux and field quantities. 
\end{itemize}

>>>>

\newcommand{\vect}[1]{\ensuremath\boldsymbol{#1}}
\newcommand{\tensor}[1]{\underline{\vect{#1}}}
\newcommand{\del}{\triangle}
\newcommand{\grad}{\nabla}
\renewcommand{\div}{\grad \cdot}
\newcommand{\ip}[1]{\left\langle #1 \right\rangle}
\newcommand{\eip}[1]{a\left( #1 \right)}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdd}[2]{\frac{\partial^2#1}{\partial#2^2}}
\newcommand{\nor}[1]{\left\| #1 \right\|}

@@

\newcommand{\Reyn}{\rm Re}
\newcommand{\Gh}{\Gamma_h}
\newcommand{\Oh}{\Omega_h}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\jump}[1] {\ensuremath{\LRs{\![#1]\!}}}

@@

\section{The compressible Navier-Stokes equations}

Following Demkowicz and Oden \cite{Demkowicz:1990:NFE:112271.112276}, we consider the compressible Navier-Stokes equations
\[
\sum^2_{i=1} F_i\left( U\right)_{,i}=\sum^2_{i=1} G_i\left( U\right)_{,i}
\]
where the Eulerian fluxes are given as
\[
F_1 = \vectfour{\rho u_1}{\rho u_1^2+p}{\rho u_1 u_2}{((\rho e)+p)u_1}, \qquad F_2 = \vectfour{\rho u_2}{\rho u_1 u_2}{\rho u_2^2+p}{((\rho e)+p)u_2},
\]
and the viscous fluxes are given as
\[
G_i = \vectfour{0}{\sigma_{i1}}{\sigma_{i2}}{\sum_{j=1}^2\sigma_{ij}u_j + q_i}.
\]
We have also the constitutive relations 
\begin{align*}
\sigma_{ij} &= \mu(u_{i,j}+u_{j,i}) + \lambda u_{k,k}\delta_{ij}\\
q_i &= \kappa T_{,i}\\
p &= (\gamma-1)\rho\iota\\
\iota &= e-\frac{1}{2}(u_1^2+u_2^2)\\
\iota &= c_vT
\end{align*}
where $e$ and $\iota$ are energy and internal energy per unit mass, respectively. We write the Navier Stokes equations as a system of seven first order equations. Let $[u_1,u_2] = [u,v]$.

@@

We note that, though this is a standard approach in solid mechanics, it is nonstandard compared to the usual finite element and DG approaches to the viscous stresses. We adopt such an approach to better mirror our experiences with the convection-diffusion equation \cite{DPGrobustness,DPGrobustness2}.

In general, $\mu$ and $\lambda$ are functions of temperature, obeying the power law
\[
\mu = \left(\frac{T}{T_0}\right)^\beta,
\]
where $T_0$ is a reference temperature. We choose $\beta = 2/3$ in this case.

\subsection{Fourier's heat conduction law}
We assume Fourier's law 
\begin{align*}
\vec{q} &= \kappa \grad T,
\end{align*}
where $\kappa$ is generally a function of temperature. We introduce here the Prandtl number here as well
\[
{\rm Pr} = \frac{\gamma c_v \mu}{\kappa}.
\]
In this case, we assume a constant Prandlt number, which implies that the heat conductivity $\kappa$ is proportional to viscosity $\mu$.

@@

We strongly enforce symmetry of the stress tensor $\boldsymbol \sigma$ by setting $\sigma_{21} = \sigma_{12}$.

\section{Nondimensionalization}
To nondimensionalize our equations, we introduce nondimensional quantities for length, density, velocity, temperature, and viscosity. 
\[
\boldsymbol x^* = \frac{\boldsymbol x}{L}, \qquad \rho^* = \frac{\rho}{\rho_{\infty}}, \qquad \boldsymbol u^* = \frac{\boldsymbol u}{V_\infty}, \qquad T^* = \frac{T}{T_\infty}, \qquad \mu^* = \frac{\mu}{\mu_\infty}
\]
Momentum, pressure, internal energy, and bulk viscosity are then nondimensionalized with respect to the above variables
\[
\boldsymbol m^* = \frac{\boldsymbol m}{\rho_\infty V_\infty}, \qquad p^* = \frac{p}{\rho_\infty V_\infty^2}, \qquad \iota^* = \frac{\iota}{V_\infty^2}, \qquad \lambda^* = \frac{\lambda}{\mu_\infty}
\]
We introduce, for convenience, the Reynolds number
\[
\Reyn = \frac{\rho_\infty V_\infty L}{\mu_\infty} 
\]
and the reference (free stream) Mach number
\[
M_\infty = \frac{V_\infty}{\sqrt{\gamma(\gamma-1)c_vT_\infty}}
\]
Note that 
\[
a = \sqrt{\frac{\gamma p_\infty}{\rho_\infty}} = \sqrt{{\gamma p_\infty}} = \sqrt{\gamma(\gamma-1)c_vT_\infty}
\]
The equations take the same form as before after nondimensionalization, so long as we define new material constants
\[
\tilde{\mu} = \frac{\mu^*}{\Reyn}, \qquad \tilde{\lambda} = \frac{\lambda^*}{\Reyn}, \qquad \tilde{c}_v = \frac{1}{\gamma(\gamma-1)M_\infty^2}, \qquad \tilde{\kappa} = \frac{\gamma\tilde{c}_v\tilde{\mu}}{\rm Pr}
\]
From here on, we drop the $^*$ superscript and assume all variables refer to their nondimensionalized quantities.

@@

We do not linearize the viscosities $\mu$ and $\lambda$, but instead set them based on the power law and the solution at the previous timestep for simplicity.

@@

For our initial guesses, we set
\[
\rho = 1,\qquad u_1= 1,\qquad u_2 = 0, \qquad T = 1
\]
which is consistent with what was done in Demkowicz and Oden in \cite{Demkowicz:1990:NFE:112271.112276}. Stresses are set uniformly to zero.

\section{Test norm}

For the convection-diffusion problem 
\begin{align*}
\div \left(\beta u - \sigma\right) &= f\\
\frac{1}{\epsilon}\sigma - \grad u &= 0
\end{align*}
on domain $\Omega$, with mesh $\Oh$ and mesh skeleton $\Gh$, the DPG variational formulation is
\begin{align*}
b\left(\left(u,\sigma, \widehat{u}, \widehat{f}_n\right),
\left( v, \tau \right)\right) = \left(u,\div \tau - \beta \cdot \grad
v\right)_{\Oh} + \left(\sigma, \epsilon^{-1} \tau + \grad v\right)_{\Oh} - \LRa{
\jump{\tau\cdot n}, \widehat{u} }_{\Gh} + \LRa{ \widehat{f}_n,
  \jump{v} }_{\Gh}.
\end{align*}
with $v\in H^1$ and $\tau \in H({\rm div},\Oh)$.

The test norm adopted for convection-diffusion in \cite{DPGrobustness,DPGrobustness2} is defined elementwise on $K$ as
\[
\|\left(v,\tau\right)\|_{V,K}^2 = \min\left\{\frac{\epsilon}{|K|},1\right\}\|v\|^2 + \epsilon \|\grad v\|^2 + \|\beta \cdot \grad v\|^2 + \| \div \tau\|^2 + \min\left\{\frac{1}{\epsilon},\frac{1}{|K|}\right\}\|\tau\|^2.
\]
This test norm both delivers robust control of the error in the $L^2$ variables $u$ and $\sigma$ and avoids boundary layers in the computation of local test functions.

Our test norm is extrapolated to the Navier-Stokes equations as follows: we denote the vector of $H^1$ test functions as $v=\{v_1,v_2,v_3,v_4\}$, and similarly for $W = \{\tau_1,\tau_2,\tau_3\}$. Similarly, we group our Eulerian and stress variables into the vector variables $U$ and $\Sigma$, respectively. If $R_{\rm Euler}(U,\Sigma)$ and $R_{\rm visc}(U,\Sigma)$ are Eulerian and viscous nonlinear residuals, our formulation for the linearized Navier-Stokes equations is
\begin{align*}
\div \left(A_{\rm Euler}U - A_{\rm visc}\Sigma\right) &= R_{\rm Euler}(U,\Sigma)\\
E_{\rm visc} \Sigma - \grad U &= R_{\rm visc}(U,\Sigma)
\end{align*}
with variational formulation
\begin{align*}
\LRa{\widehat{F}_n,V}_{\Gh} + \left(U,\div W - A_{\rm Euler}^T\grad  V\right) + \LRa{\widehat{U},W\cdot n}_{\Gh} + \left(\Sigma,E_{\rm visc}^TW - A_{\rm visc}^T\grad  V\right) &= 0
\end{align*}
Identifying vector-valued terms in the Navier-Stokes formulation with equivalent scalar terms in the convection-diffusion equation allows us to extrapolate our test norm to systems of equations
\[
\|\left(V,W\right)\|_{V,K}^2 = \min\left\{\frac{\Reyn}{|K|},1\right\}\|v\|^2 + \frac{1}{\Reyn} \|A_{\rm visc}^T\grad v\|^2 + \|A_{\rm Euler}^T \grad v\|^2 + \| \div W\|^2 + \min\left\{\Reyn,\frac{1}{|K|}\right\}\|E_{\rm visc}^T\tau\|^2.
\]
An advantage of this extrapolation approach is that the incompletely parabolic nature of the Navier-Stokes equation is taken into account; there is no diffusive term present in the mass conservation equation, and the test norm reflects that by requesting only limited regularity of $v_1$.\footnote{The situation is analagous to using the full $H^1(\Oh)$ norm for the pure convection equation --- the optimal test norm $\nor{v}_V = \nor{\beta\cdot \grad v} + \nor{v}$ implies only streamline regularity, whereas taking $\nor{v}_V = \nor{\grad v} + \nor{v}$ implies stronger regularity on the test space $V$ than the graph norm. Consequently, convergence is suboptimal for DPG applied to the convection problem under the $H^1(\Oh)$ test norm.}

\section{Boundary conditions}

As a consequence of the ultra-weak variational formulation, our solution is linear in the flux and trace variables. Thus, the nonlinear boundary conditions can be applied directly to our fluxes $\widehat{f}_{i,n}, i = 1,\ldots,4$, and traces $\widehat{u}_1$, $\widehat{u}_2$, and $\widehat{T}$.

Additionally, inflow boundary conditions are applied not directly to the trace variables $\widehat{u}_1$, $\widehat{u}_2$, and $\widehat{T}$, but to the fluxes $\widehat{f}_{i,n}$. Extrapolating from the convection-diffusion problem, this allows us to use a stronger test norm without experiencing adverse effects for smaller diffusion/higher Reynolds numbers \cite{DPGrobustness,DPGrobustness2}.

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=.5]{figs/flat_plate_BCs.pdf}
\caption{Carter flat plate problem.}
\end{figure}

\begin{itemize}
\item \textbf{Inflow boundary conditions:} free stream conditions are applied here to all four fluxes $\widehat{f}_{i,n}$.
\item \textbf{Symmetry boundary conditions:} $u_n = q_n = \pd{u_s}{n} = 0$. Here, this implies $u_2 = q_2 = \sigma_{12} = 0$. We impose the stress condition by noting that, for the flat plate geometry, if $u_2 = 0$, then at the top and bottom, with $n = (0,1)$, $\widehat{f}_{2,n} = \sigma_{12}$, and $\widehat{f}_{4,n} = q_2$ if $\sigma_{12}$ and $u_2 = 0$. 
\item \textbf{Flat plate boundary conditions:} $u_1 = u_2 = 0$, and $T = T_w = \left[1+(\gamma-1)M_\infty^2/2\right] T_\infty = 2.8T_\infty$ (for Mach 3 flow). We impose these strongly on the trace variables $\widehat{u}_1, \widehat{u}_2, \widehat{T}$. 
\item \textbf{Outflow boundary conditions:} the exact boundary conditions to enforce here are not universally agreed on\footnote{Demkowicz et al enforce an outflow boundary condition only in regions where the flow is subsonic.\cite{Demkowicz:1990:NFE:112271.112276}}. Many enforce $\pd{u_1}{n}=\pd{u_2}{n}=0$ and $\pd{T}{n} = 0$. We adopt a ``no boundary condition'' outflow condition, first introduced in \cite{FLD:FLD1650140506}. A mathematical analysis and explanation of this boundary condition for standard $H^1$ elements is given in \cite{FLD:FLD505}.

An alternative boundary condition would be as follows: since $u_1$ and $u_2$ are not expected to be zero at these points, we can set $\widehat{f}_{2,n}$ and $\widehat{f}_{4,n}$ to their represented quantities using either the field variables from the background flow, or (in the case of pseudo-time stepping) the previous timestep's flux and field quantities.

\end{itemize}

\section{Numerical experiments}

For our solver, we use a standard pseudo-time solver and a standard greedy refinement scheme. Though we have not yet implemented adaptive timestepping, we are able to take large enough uniform timesteps over a coarse mesh such that convergence is comparable to the convergence of a damped Newton method.

The parameters used are as follows:
\begin{itemize}
\item{DPG parameters:} Both $p$ and $\Delta p = 2$ uniformly across the mesh. 
\item{Adaptivity parameters:} Energy threshold for refinements is $\alpha = .2$ or $\alpha = .15$. 
\item{Time-stepping parameters:} $\Delta t = .1$, and tolerance for transient residual $\epsilon_t = 5e-7$.
\end{itemize}

\subsection{Carter plate problem}

For the flat plate problem, we take the domain $\Omega = [0,2]\times[0,1]$. For $\Reyn = 500$, we begin with an initial mesh of 8 by 16 elements. 
\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=.5]{figs/rhoRe500.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/u1Re500.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/u2Re500.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/TRe500.pdf}}
\caption{Solutions after four refinements for $\rho$, $u_1$, $u_2$, and $T$ and $\Reyn = 500$.}
\label{fig:Re500}
\end{figure}

For $\Reyn = 1000$, we experimented with both an initial mesh of 4 by 8 and 8 by 16 elements. We note that, on a very coarse initial mesh (4 by 8 elements), the convergence of the pseudo-timestep stagnates at around a transient residual of $1e-6$; however, if we lower the tolerance only for the first step and continued with the adaptive algorithm, the solution converges below $\epsilon_t$ for all refined meshes. 
\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=.5]{figs/rhoRe1000.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/u1Re1000.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/u2Re1000.pdf}}
\subfigure{\includegraphics[scale=.5]{figs/TRe1000.pdf}}
\caption{Solutions after four refinements for $\rho$, $u_1$, $u_2$, and $T$ and $\Reyn = 500$.}
\label{fig:Re500}
\end{figure}

\begin{figure}[!h]
\centering
\subfigure[$\Reyn = 500$]{\includegraphics[scale=.5]{figs/MeshRe500.pdf}}
\subfigure[$\Reyn = 1000$]{\includegraphics[scale=.5]{figs/MeshRe1000.pdf}}
\caption{Adaptive mesh after four refinement iterations for $\Reyn = 500$ and $\Reyn = 1000$.}
\label{fig:MeshRe500}
\end{figure}

\bibliographystyle{plain}
\bibliography{notes}
